A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-08-29 masters_thesis/si_paper/paper.tex
Made modifications based on editor in chief's comments

Particle swarm optimization (PSO) has previously been parallelized primarily by
distributing the computation corresponding to particles across multiple
processors.  In these approaches, adding processors can only increase the swarm
size.  However, in many cases this is not efficient when scaled to very large
swarm sizes (on very large clusters).  Current methods cannot answer well the
question, ``How can 1000 processors be fully utilized when 50 or 100 particles
is the most efficient swarm size?''  In this paper we attempt to answer that
question with a speculative approach to the parallelization of PSO that we
refer to as SEPSO.

In our approach, we refactor PSO such that the computation needed for iteration
$t+1$ can be done concurrently with the computation needed for iteration $t$.
Thus we can perform two iterations of PSO at once.  Even with some amount of
wasted computation, we show that this approach to parallelization in PSO often
outperforms the standard parallelization of simply adding particles to the
swarm.  SEPSO produces results that are exactly equivalent to PSO; this is not
a new algorithm or variant, only a new method of parallelization.

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization
problems~\citep{poli-2008-pso-applications}.  For much of the algorithm's
history, PSO was run serially on a single machine.  However, the world's
computing power is increasingly coming from large clusters of processors.  In
order to efficiently utilize these resources for computationally intensive
problems, PSO needs to run in parallel.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of function evaluations performed.  Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm, and \sec{related} gives a brief overview
of previous parallelization techniques for this algorithm.  \sec{sepso} shows
mathematically how speculative evaluation can be done in parallel PSO to
perform two iterations at once, leaving implementation concerns to the
appendices.  In \sec{relax}, we discuss various methods of improving the
performance of speculative evaluation in PSO, all of which break the
requirement of strictly reproducing the behavior of the original algorithm.
\sec{setup} describes the experiments we ran, and \sec{results} presents the
results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

The motion of particles through the search space has three components: an
inertial component that gives particles momentum as they move, a cognitive
component where particles remember the best solution they have found and are
attracted back to that place, and a social component by which particles are
attracted to the best solution that any of their neighbors have found.

@@

The idea of speculative decomposition in the parallelization literature is not
new \citep{grama-2003-intro-to-parallel-computing}.  In the field of function
optimization, simulated annealing has previously been parallelized using this
technique~\citep{witte-1991-parallel-simulated-annealing-speculative}, though
we are not aware of other evolutionary or swarm-intelligence based algorithms
having been parallelized with speculative decomposition.

@@

There are many ways to parallelize the basic PSO algorithm.  The most
fundamental decision to make in parallel PSO is which parallel architecture to
use.  Several architectures have been proposed, including Master-Slave, fully
distributed (sometimes called ``diffusion''), and reformulating PSO into
Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
mcnabb-2007-parallel-pso-using-mapreduce}.  A somewhat orthogonal
implementation decision when parallelizing PSO is whether to have synchronous
communication or asynchronous communication.

Synchronous parallel implementations of PSO reproduce the standard serial
algorithm exactly.  This approach was first described analytically by
\citet{belal-2004-parallel-models-for-pso} and first implemented by
\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
master-slave algorithm, the master assigns tasks to slave processors, and in
parallel PSO, each task consists primarily of a function evaluation.  Updating
the particle's position and value may also be included in the
task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
in serial by the
master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
proceeding to the next iteration, particles communicate, and each particle
updates its neighborhood best.  Whether this communication step happens
sequentially on the master or in parallel, each particle must receive
communication from its neighbors before proceeding.  The benefits of the
synchronous PSO include its simplicity, repeatability, and comparability with
standard PSO, which may be essential in research applications.

@@

The other area of research in parallelizing PSO deals not with the
implementation details of architecture and synchronicity, but with what should
be done with the PSO equations when many hundreds or thousands of processors
are available.  The main issues that have been addressed in this space are how
many particles to use for a particular number of processors and what
communication topology should be employed.

@@

With regard to increasing the swarm size in PSO, some recent work has suggested
that increasing the swarm size throughout the course of the optimization
process provides better results than having a set swarm
size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
focused on serial computation and are based on total number of function
evaluations, which, when running in parallel on expensive functions, is less
important than total number of iterations.  Other work focusing on
parallelization has shown that when extra processors are available they should
be used, as performance increases with swarm size when measuring in terms of
number of iterations~\citep{mcnabb-2009-large-particle-swarms,
jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
course of the optimization process, some processors would be sitting idle at
most iterations.

The contribution of our work is in this area of what should be done with the
PSO equations to better utilize a thousand processors when they are available.
In our work we use a synchronous, MapReduce implementation of parallel PSO.
While we use a specific implementation, we describe how speculative evaluation
can be performed in any of the synchronous architectures mentioned in the
previous section.  The adaptation of our methods to asynchronous PSO
parallelization methods should be straightforward, though it is left to future
work.

@@

The refactoring hinges on the idea that there are only a few possible new
positions, or updates, for $\nbest$ and $\pbest$ (assuming the random
coefficients $\pURand_{t}$ and $\nURand_{t}$ have been drawn).  For the Ring
topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors find a position that updated its neighborhood
best.  In the equations that follow, we refer to an update case as $\casegen$,
and to the set of cases collectively as $\caseset$.

@@

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then Sections~\ref{sec:pruning}
through~\ref{sec:wrong} present a technique that reduces the number of
speculative evaluations that need to be done for each particle, allowing
speculative evaluation to use larger swarm sizes with the same number of
processors.  Finally, \sec{manyiters} shows a method for speculating several
iterations ahead, instead of just one.

@@

We frequently present tables summarizing our results.  In each table, we bold
the ``best'' method, meaning that it has at least a 90\% success rate and its
average time to completion is faster than all other methods that have at least
a 90\% success rate\footnote{``Success rate'' here means that the algorithm
reached some predefined value which is mentioned in the table or figure
wherever a success rate is reported.}.  We used a double-sided t-test to test
for statistical significance in these results, and all bolded values are
statistically significant with a p-value of less than $10^{-5}$.  In some
instances two values are bolded because they are statistically tied, while both
are significantly better than all other methods.

@@

We mentioned related work showing that increasing the swarm size throughout the
course of the algorithm could provide improved performance over a fixed swarm
size in serial PSO~\citep{montes-de-oca-2010-incremental-social-learning-pso}.
If this method were extended to parallel PSO, most processors would be idle in
the first few iterations, while more would be utilized at the end.  During
iterations where there are many un-utilized processors, a natural use of them
would be speculative evaluation, performing two or more of those iterations at
a time.

>>>>

Particle swarm optimization (PSO) has previously been parallelized primarily by
distributing the computation corresponding to particles across multiple
processors.  In these approaches, the only benefit of additional processors is
an increased swarm size.  However, in many cases this is not efficient when
scaled to very large swarm sizes (on very large clusters).  Current methods
cannot answer well the question: ``How can 1000 processors be fully utilized
when 50 or 100 particles is the most efficient swarm size?''  In this paper we
attempt to answer that question with a speculative approach to the
parallelization of PSO that we refer to as SEPSO.

In our approach, we refactor PSO such that the computation needed for iteration
$t+1$ can be done concurrently with the computation needed for iteration $t$.
Thus we can perform two iterations of PSO at once.  Even with some amount of
wasted computation, we show that this approach to parallelization in PSO often
outperforms the standard parallelization of simply adding particles to the
swarm.  SEPSO produces results that are exactly equivalent to PSO; that is,
SEPSO is a new method of parallelization and not a new PSO algorithm or
variant.

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization
problems~\citep{poli-2008-pso-applications,
poli-2007-particle-swarm-optimization}.  For much of the algorithm's history,
PSO was run serially on a single machine.  However, the world's computing power
is increasingly coming from large clusters of processors.  In order to
efficiently utilize these resources for computationally intensive problems, PSO
needs to run in parallel.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, showing performance after
  a set number of function evaluations.  Error bars show mean and 10th and 90th
  percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

This paper is organized as follows. \sec{pso} describes the particle swarm
optimization algorithm, and \sec{related} gives a brief overview of previous
parallelization techniques for this algorithm.  \sec{sepso} shows
mathematically how speculative evaluation can be done in parallel PSO to
perform two iterations at once, leaving implementation concerns to the
appendices.  In \sec{relax}, we discuss various methods of improving the
performance of speculative evaluation in PSO, all of which break the
requirement of strictly reproducing the behavior of the original algorithm.
\sec{setup} describes the experiments we ran, and \sec{results} presents the
results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

The motion of particles through the search space has three components: an
inertial component that gives particles momentum as they move, a cognitive
component where particles remember the best solution they have found and are
attracted back to that place, and a social component by which particles are
attracted to the best solution that any of their neighbors have found.  We use
the constricted PSO algorithm, which was recommended as the standard for PSO
research~\citep{bratton-2007-defining-a-standard-for-pso}.

@@

The idea of speculative decomposition in the parallelization literature is not
new \citep{grama-2003-intro-to-parallel-computing}.  In the field of function
optimization, simulated annealing has previously been parallelized using this
technique~\citep{witte-1991-parallel-simulated-annealing-speculative}, though
we are not aware of other evolutionary or swarm intelligence based algorithms
having been parallelized with speculative decomposition.

@@

There are many ways to parallelize the basic PSO algorithm.  The most
fundamental decision to make in parallel PSO is which parallel architecture to
use.  Several architectures have been proposed, including Master-Slave, fully
distributed (sometimes called ``diffusion''), and reformulating PSO into
Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
mcnabb-2007-parallel-pso-using-mapreduce}.

An important part of the parallel architecture is the type of interprocessor
communication used.  PSO has previously been parallelized with both synchronous
and asynchronous communication.  Synchronous parallel implementations of PSO
reproduce the standard serial algorithm exactly.  This approach was first
described analytically by \citet{belal-2004-parallel-models-for-pso} and first
implemented by \citet{schutte-2004-parallel-global-optimization-with-pso}.  In
a typical master-slave algorithm, the master assigns tasks to slave processors,
and in parallel PSO, each task consists primarily of a function evaluation.
Updating the particle's position and value may also be included in the
task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
in serial by the
master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
proceeding to the next iteration, particles communicate, and each particle
updates its neighborhood best.  Whether this communication step happens
sequentially on the master or in parallel, each particle must receive
communication from its neighbors before proceeding.  The benefits of the
synchronous PSO include its simplicity, repeatability, and comparability with
standard PSO, which may be essential in research applications.

@@

The other area of research in parallelizing PSO deals not with the
implementation details of architecture and synchronicity, but with what should
be done with the PSO equations when many hundreds or thousands of processors
are available.  The main issues that have been addressed are how many particles
to use for a particular number of processors and what communication topology
should be employed.

@@

With regard to increasing the swarm size in PSO, some recent work has suggested
that increasing the swarm size throughout the course of the optimization
process provides better results than having a set swarm
size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
montes-de-oca-2011-incremental-social-learning-pso}.  However, these results
focused on serial computation and are based on total number of function
evaluations, which, when running in parallel on expensive functions, is less
important than total number of iterations.  Other work focusing on
parallelization has shown that when extra processors are available they should
be used, as performance increases with swarm size when measuring in terms of
number of iterations~\citep{mcnabb-2009-large-particle-swarms,
jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
course of the optimization process, some processors would be sitting idle at
most iterations.

The contribution of our work lies in the realm of efficiently utilizing
parallel resources.  In our work we use a synchronous, MapReduce implementation
of parallel PSO.  While we use a specific implementation, we describe how
speculative evaluation can be performed in any of the synchronous architectures
mentioned in the previous section.  The adaptation of our methods to
asynchronous PSO parallelization methods should be straightforward, though it
is left to future work.

@@

The refactoring hinges on the idea that there are only a few possible new
positions, or updates, for $\pbest$ and $\nbest$ (assuming the random
coefficients $\pURand_{t}$ and $\nURand_{t}$ have been drawn).  For the Ring
topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors find a position that updated its neighborhood
best.  In the equations that follow, we refer to an update case as $\casegen$,
and to the set of cases collectively as $\caseset$.

@@

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then Section~\ref{sec:pruning} presents a
technique that reduces the number of speculative evaluations that need to be
done for each particle, allowing speculative evaluation to use larger swarm
sizes with the same number of processors.  Finally, \sec{manyiters} shows a
method for speculating several iterations ahead, instead of just one.

@@

We frequently present tables summarizing our results.  In each table, we bold
the ``best'' method, meaning that it has at least a 90\% success rate and its
average time to completion is faster than all other methods that have at least
a 90\% success rate\footnote{``Success rate'' here means that the algorithm
reached some predefined value which is mentioned in the table or figure
wherever a success rate is reported.}.  We used a double-sided t-test to test
for statistical significance in these results, and all algorithms whose values
are bolded are significantly superior to non-bolded algorithms with a p-value
of less than $10^{-5}$.

@@

We mentioned related work showing that increasing the swarm size throughout the
course of the algorithm could provide improved performance over a fixed swarm
size in serial PSO~\citep{montes-de-oca-2011-incremental-social-learning-pso}.
If this method were extended to parallel PSO, most processors would be idle in
the first few iterations, while more would be utilized at the end.  During
iterations where there are many un-utilized processors, a natural use of them
would be speculative evaluation, performing two or more of those iterations at
a time.
